---
title: 'Mini-Project: COVID-19 Vaccination Rates'
author: "Divya Shetty (A15390408)"
date: "3/3/2022"
output:
  pdf_document:
    latex_engine: xelatex
---


## Getting Started

Import and examine the vaccination data.

```{r message=FALSE, warning=FALSE}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

**Q1. What column details the total number of people fully vaccinated?** 

The column is called "persons_fully_vaccinated".


**Q2. What column details the Zip code tabulation area?** 

The column is "zip_code_tabulation_area".


**Q3. What is the earliest date in this dataset?** 

```{r}
vax$as_of_date[1]
```
The earliest date is "2021-01-05".

**Q4. What is the latest date in this dataset?**

```{r}
vax$as_of_date[length(vax$as_of_date)]
```
The latest date is "2022-03-01".



Use the skim() function to get an overview of the data.

```{r}
skimr::skim(vax)
```


**Q5. How many numeric columns are in this dataset?**

There are 9 numeric columns.

**Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?**

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

There are 18338 NA values for "persons_fully_vaccinated".

**Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?**

```{r}
sum(is.na(vax$persons_fully_vaccinated)) / length(vax$persons_fully_vaccinated)
```

About 17.04% of the values in "persons_fully_vaccinated" are missing.

**Q8. [Optional]: Why might this data be missing?**

Missing values may be due to no records being collected from a given county.


## Working with Dates

The lubridate package let's us use date data in a useful manner. Convert dates into lubridate formate and perform math operations with it!

```{r message=FALSE}
library(lubridate)
today()

#specify the year-month-day format when converting
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
#how many days have passed since the first vaccination?
today() - vax$as_of_date[1]
```

```{r}
#how many days does the dataset span?
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```


**Q9. How many days have passed since the last update of the dataset?**

```{r}
today() - vax$as_of_date[nrow(vax)]
```

5 days have passed since the last update of the dataset.

**Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?**

```{r}
length(unique(vax$as_of_date))
```

There are 61 different dates in the dataset.


## Working with ZIP Codes

In R, we can use the zipcodeR package to make working with ZIP codes - a postal code used by the United States Postal Service (USPS) - easier.

```{r}
library(zipcodeR)

#calculate the distance between the centroids of two zip codes in miles
zip_distance('92037','92109')

#pull census data for zip code areas
reverse_zipcode(c('92037', "92109") )
```


## Focus on the San Diego Area

Use the dplyr package to restrict the data to only the San Diego area.

```{r message=FALSE}
library(dplyr)

sd <- filter(vax, county == "San Diego")
nrow(sd)
```

The dplyr package can also be useful when trying to subset with multiple criteria. For example:

```{r}
sd.10 <- filter(vax, county == "San Diego" & age5_plus_population > 10000)
```


**Q11. How many distinct zip codes are listed for San Diego County?**

```{r}
length(unique(sd$zip_code_tabulation_area))
```

There are 107 unique zip codes for the San Diego County.

**Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?**

```{r}
sd$zip_code_tabulation_area[which.max(sd$age12_plus_population)]
```

The '92154' zip code area has the largest 12+ population.


Select all San Diego “county” entries on “as_of_date” “2022-02-22” and use this for the following questions.

```{r}
sd.feb <- filter(vax, county == "San Diego" & as_of_date == "2022-02-22")
```

**Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”?**

```{r}
mean(sd.feb$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

The average percent of population fully vaccinated in the San Diego County as of 2022-02-22 is 70.42%.

**Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”?**

```{r warning=FALSE}
library(ggplot2)

ggplot(sd.feb) +
  aes(x = percent_of_population_fully_vaccinated) +
  geom_histogram(bins = 12) +
  labs(title = "Histogram of Vaccination Rates Across San Diego County", 
       subtitle = "As of 2022-02-22", 
       x = "Percent of Population Fully Vaccinated in Zip Code Area", 
       y = "Count (Zip Code Areas)")
```


## Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144. Use this information to subset the data and check that it's done correctly.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")
ucsd[1,]$age5_plus_population
```

**Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area.**

```{r}
ggplot(ucsd) +
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group = 1) +
  ylim(c(0, 1)) +
  labs(title = "Vaccination Rate for La Jolla CA 92037", 
       x = "Date", y = "Percent Vaccinated")
```


## Comparing to Similar Size Areas

Look across every zip code area with a population at least as large as that of 92037 on as_of_date “2022-02-22”.

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 & as_of_date == "2022-02-22")

head(vax.36)
```

**Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?**

```{r}
avg.per <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm = TRUE)
avg.per
```

```{r}
ggplot(ucsd) +
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group = 1) +
  ylim(c(0, 1)) +
  labs(title = "Vaccination Rate for La Jolla CA 92037", 
       x = "Date", y = "Percent Vaccinated") +
  geom_hline(aes(yintercept = avg.per, color = "red"), 
             linetype = "dashed", show.legend = FALSE)
```


**Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?**

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```


**Q18. Using ggplot generate a histogram of this data.**

```{r}
ggplot(vax.36) +
  aes(x = percent_of_population_fully_vaccinated) +
  geom_histogram(bins = 30) +
  labs(x = "Percent Vaccinated", y = "Count") +
  xlim(c(0, 1))
```


**Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?**

```{r}
#92109
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
#92040
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area == "92040") %>%
  select(percent_of_population_fully_vaccinated)
```

Both zip code areas are below the average value calculated.
 
**Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with age5_plus_population > 36144.**

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(x = as_of_date,
      y = percent_of_population_fully_vaccinated, 
      group = zip_code_tabulation_area) +
  geom_line(alpha = 0.2, color = "blue") +
  ylim(c(0, 1)) + 
  labs(x = "Date", y = "Percent Vaccinated",
       title = "Vaccination Rate Across California",
       subtitle = "Only areas with a population above 36k are shown") +
  geom_hline(yintercept = avg.per, linetype = "dashed")
```


**Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?**

I would not feel comfortable traveling for break considering that nearly half of the largely populated areas don't have a great majority of their population vaccinated. This would make having in-person classes after break riskier since we can't guarantee that each location everyone traveled to was safe.