---
title: 'Week 6: Machine Learning with PCA'
author: "Divya Shetty (A15390408)"
date: "2/13/2022"
output:
  pdf_document: default
  html_document: default
---

# Principal Component Analysis [PCA]

## PCA of UK Food Data

The UK food data is imported from a given input csv file.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)

x
```
**Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?**
```{r}
dim(x)
```
There are 17 rows and 5 columns in data frame 'x'. You can use dim(x) or nrow(x) and ncol(x) to determine this.

<br>

There should be 4 columns: one for each country. Data frame is adjusted accordingly.

```{r}
x <- read.csv(url, row.names = 1)

x
```
**Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?**  
I prefer adjusting the row names in the read.csv() function since it requires less lines of code. It's also more robust because running "x <- x[,-1])" multiple times will continuously adjust the row names until there are no more columns left whereas editing read.csv() doesn't have this problem.

<br>

A few types of data visualizations are attempted, though they are uninformative.

```{r}
#barplot visualization
barplot(as.matrix(x), beside = TRUE, col = rainbow(nrow(x)))
```

**Q3: Changing what optional argument in the above barplot() function results in the following plot?**  
Change the value of beside to FALSE. This argument indicates whether the bar plot should have the columns side-by-side (TRUE) or have the stacked (FALSE)
```{r}
barplot(as.matrix(x), beside = FALSE, col = rainbow(nrow(x)))
```

<br>

**Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?**
```{r}
#pairwise plot visualization
pairs(x, col = rainbow(10), pch = 16)
```

Each plot illustrates the data comparison between two given countries. A given point on the diagonal of a given plot means the same amount of some food is consumed between the two countries.

**Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?**  
N. Ireland appears to have more differences in food consumptions than the other countries compared to one another as shown by the N. Ireland plots having more points spread out.

<br>

PCA to the rescue! PCA helps with analysis of many variables or large datasets.
The main base R function is "prcomp()". This functions expects rows are observations and columns are variables, so input must be transposed.

```{r}
pca <- prcomp(t(x))

#extra info about pca
attributes(pca)
summary(pca)
```


To create our PCA plot, we must access the "pca$x" component.

**Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.**
```{r}
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2", xlim = c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))
```

**Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.**
```{r}
country_color <- c("orange", "red", "blue", "green")

plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2", xlim = c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col = country_color)
```


The square of "pca$sdev" can be used to calculate how much variation in the original data each PC accounts for.

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
barplot(v, xlab = "Principal Component", ylab = "Percent Variation")
```

<br>

Digging deeper! The "pca$rotation" component can determine the influence of each of the original variables on the principal components ie. the loading scores.

```{r}
#loadings plot for PC1
par(mar = c(10, 3, 0.35, 0))
barplot(pca$rotation[,1], las = 2 )
```

**Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominently and what does PC2 maninly tell us about?**
```{r}
#loadings plot for PC2
par(mar = c(10, 3, 0.35, 0))
barplot(pca$rotation[,2], las = 2 )
```

Fresh_potatoes and Soft_drinks are both featured prominently in the plot for PC2. Since Fresh_potatoes has a positive score this indicates that this food "pushes countries upwards, and since Soft_drinks has a negative score, it "pushes" countries downwards in the PCA plot.

<br>

Biplots can also be used to visualize this information, usually for small datasets.
```{r}
biplot(pca)
```

<br>

## PCA of RNA-Seq Data

Read the RNA-seq count dataset into the "rna.data" data frame.

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names = 1)
head(rna.data)
```

**Q10: How many genes and samples are in this data set?**
```{r}
dim(rna.data)
```

There are 10 genes and 100 samples.

<br>

A PCA plot is generated for the above data. Remember, the data must be transposed! A summary is generated to show how much variation in the original data each PC accounts for.

```{r}
pca2 <- prcomp(t(rna.data), scale = TRUE)
plot(pca2$x[,1], pca2$x[,2], xlab = "PC1", ylab = "PC2")
```
```{r}
#variation info
summary(pca2)
```


Using the information provided from the summary, a scree plot can be generated to visualize the how much variation in the original data each PC accounts for.

```{r}
#variance per PC
pca2.var <- pca2$sdev^2

#percent variance bc more informative to look at 
pca2.var.per <- round(pca2.var/sum(pca2.var)*100, 1)

#make scree plot
barplot(pca2.var.per, main = "Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab = "Principal Component", ylab = "Percent Variation")
```


Now! Make the PCA plot look more presentable.

```{r}
#designate colors
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

#make PCA plot
plot(pca2$x[,1], pca2$x[,2], col = colvec, pch = 16,
     xlab = paste0("PC1 (", pca2.var.per[1], "%)"),
     ylab = paste0("PC2 (", pca2.var.per[2], "%)"))

#labels
text(pca2$x[,1], pca2$x[,2], labels = colnames(rna.data), pos = c(rep(4,5), rep(2,5)))
```


We can also use ggplot2 to visualize the data in a different format.

```{r}
library(ggplot2)

#convert PCA data into data frame with 'wt' and 'ko'
df <- as.data.frame(pca2$x)
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

#ggplot visualization
ggplot(df) + 
    aes(PC1, PC2, label = samples, col = condition) + 
    geom_label(show.legend = FALSE) +
    labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x = paste0("PC1 (", pca2.var.per[1], "%)"),
       y = paste0("PC2 (", pca2.var.per[2], "%)"),
       caption="BIMM143 example RNA-seq data") +
    theme_bw()
```

<br>

OPTIONAL: Finding the top ten genes that contribute the most to PC1

```{r}
loading_scores <- pca2$rotation[,1]

#greatest to least abs val to find top 10 contributors
#that's positive (+) OR negative (-)
gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing = TRUE)

#show the names of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes 
```

