---
title: 'Class 10: Population Analysis'
author: "Divya Shetty (A15390408)"
date: "2/17/2022"
output:
  pdf_document: default
  html_document: default
---

**Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.**

```{r}
#read text file
exp <- read.table("rs8067378_ENSG00000172057.6.txt")
head(exp)
```

*What are the sample sizes for the different genotypes?*
```{r}
table(exp$geno)
```

There are 108 samples for A/A, 233 samples for A/G, and 121 samples for G/G.


*What are the median expression levels for the different genotypes?*

```{r}
#genotype: A/A
summary(exp[exp$geno == "A/A", "exp"])
```
```{r}
#genotype: A/G
summary(exp[exp$geno == "A/G", "exp"])
```
```{r}
#genotype: G/G
summary(exp[exp$geno == "G/G", "exp"])
```

The median expression level for A/A is 31.25 and the mean is 31.82. The median expression level for A/G is 25.065 and the mean is 25.397. The median expression level for G/G is 20.074 and the mean is 20.0594.


**Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?**

```{r}
library(ggplot2)

#make a boxplot
ggplot(exp, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot(notch = TRUE)
```

There is a noticeable difference in the median level of expression of ORMDL3 between A/A and G/G, where the median expression level for A/A is higher than that of G/G. This would indicate that the SNP is associated with the expression of ORMDL3.